# Exercise 6

*Analysis of longitudinal data*

```{r}
date()
```

# Analysis

In this last analysis exercise, I will analyze two data sets, BPRS and RATS. The information about these data sets can be found in the book we used in our course, MABS4IODS (BPRS in Chapter 8 and RATS in Chapter 9). 

Let's load some packages:
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Task 1

Since I must SWAP the data sets, I will start implementing the analyses of Chapter 8 using the RATS data. In this dataset, three groups of rats were put on different diets, where each animal’s body weight was recorded repeatedly (approximately weekly) over a 9-week period. 

```{r}
RATSL_my <- read.csv("RATSL_data.csv")
str(RATSL_my)
```

As we can see from the dataset structure, we need to factor variables ID and Group again. 

```{r}
RATSL_my$ID <- factor(RATSL_my$ID)
RATSL_my$Group <- factor(RATSL_my$Group)
str(RATSL_my)
dim(RATSL_my)

```

The next plot show a graphic overview of the data by group of rats.

```{r}
p0 <- ggplot(RATSL_my, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) + theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL_my$Weight), max(RATSL_my$Weight)))
print(p0)

```

From the plot we can see that the rats in the three groups gained weight during the time of the stud but the rats in the group 1 started with lower weight the the others. Also, the rats on the group 3 started the study with some grams more the the rats in the group 2. Additionally, we can note that three rats show different patterns of weight if compared with the majority of the rats. 

In order to see better the  the tracking phenomenon, where the individuals that have higher values at the beginning tend to have higher values throughout the study, I will standardize the variable Weight, show the data and then draw the plot again.

```{r}
RATSL_my <- RATSL_my %>%
  group_by(Time) %>%
  mutate(stdWeight = (Weight-mean(Weight))/sd(Weight)) %>%
  ungroup()

glimpse(RATSL_my)

p1 <- ggplot(RATSL_my, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) + theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL_my$stdWeight), max(RATSL_my$stdWeight)))
print(p1)

```

As we can see in the previous graphic, the standardized data do not confirm the tracking phenomenon existence in this study.   

In order to transform the repeated measurements made on each individual into a single value, I will apply the summary measure method. First, I will create the number of work days (Time), the Summary data with mean and standard error of Weight by Group and show the results.


```{r}
n <- RATSL_my$Time %>% unique() %>% length()
RATS_myS <- RATSL_my %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()
glimpse(RATS_myS)

```

Now, let's see the plot of the mean profiles.

```{r}
p2 <- ggplot(RATS_myS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
print(p2)

```

The plot shows that there mean profiles of the three groups are not overlapping, suggesting that there is  difference between them regarding the mean Weight values.

Let's check now the summary data by Group and ID, having mean as the summary variable.

```{r}
RATSL_my8S <- RATSL_my %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()
glimpse(RATSL_my8S)
p3 <- ggplot(RATSL_my8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight)")
print(p3)

```

The plot suggests the presence of 3 possible “outliers” given that their weight values are well beyond the confidence interval.

I will create than a new data by filtering the outliers and draw the plot again.

```{r}
RATSL_my8S1 <- RATSL_my8S %>%
  filter( ID!=2 & ID!= 12 & ID!= 13)
p4 <- ggplot(RATSL_my8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight)")
print(p4)

```

As a consequence of removing the outliers the confidence intervals are tighter around the mean, confirming that the outliers were distorting the distribution of the data.

To assess any difference between the three groups, I will apply a t-test and calculate a confidence interval for the differences. I will use the data without the outliers.

First, only groups 2 and 3:

```{r}
a <- filter(RATSL_my8S1, Group != 1) 
t.test(mean ~ Group, data = a, var.equal = TRUE)

```

Now groups 1 and 3, and 1 and 2

```{r}
b <- filter(RATSL_my8S1, Group != 2)
t.test(mean ~ Group, data = b, var.equal = TRUE)

c <- filter(RATSL_my8S1, Group != 3) # Only groups 1 and 2
t.test(mean ~ Group, data = c, var.equal = TRUE)

```

The results shows differences between the groups with very small p-value for all comparisons. Also, the confidence interval of 95% is wide

Closing the analysis of the RATS dataset, I will add the variable WD1 from the original data as a new variable to the summary data and fit the linear model with the mean as the response, computing the analysis of variance table for the fitted model with ANOVA.

```{r}
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')
str(RATS)
RATSL_my8S2 <- RATSL_my8S %>%
  mutate(baseline = RATS$WD1)
fit <- lm(baseline ~ mean, data = RATSL_my8S2)
anova(fit)

```


## Task 2

In this task I will implement the analyses of Chapter 9 using the BPRS data. In this dataset, 40 male were randomly assigned to one of two treatment groups, receiving a rate on the brief psychiatric rating scale (BPRS) measured before treatment began (week 0) and then at weekly intervals for eight weeks.

Let's read the BPRS and actor variables treatment and subject:

```{r}
BPRSL_my <- read.csv("BPRSL_data.csv")
BPRSL_my$treatment <- factor(BPRSL_my$treatment)
BPRSL_my$subject <- factor(BPRSL_my$subject)
str(BPRSL_my)

```

Now, a plot will be performed to show the data:

```{r}
p5 <- ggplot(BPRSL_my, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL_my$bprs), max(BPRSL_my$bprs)))
print(p5)

```

The rate of bprs values seems to be decreasing in both treatment. Also both treatment show variation in that rate along the weeks but in treatment 2 that variation seems to be wider.

In this task, Two examples of linear mixed effects models (the random intercept model and the random intercept and slope model) will be performed. 

I will start creating a regression model and printing out a summary of the model.

```{r}
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL_my)
summary(BPRS_reg)

```

Now will create a random intercept model and print its summary.

```{r}

library(lme4)
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL_my, REML = FALSE)
summary(BPRS_ref)

```

Here I create a random intercept and random slope model, showing its summary and performing an ANOVA test on the two models.

```{r}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL_my, REML = FALSE)
summary(BPRS_ref1)
anova(BPRS_ref1, BPRS_ref)

```


Let's try now a random intercept and random slope model with the interaction

```{r}
BPRS_ref2 <- lmer(bprs ~ week + treatment + (week | subject) + week * treatment, data = BPRSL_my, REML = FALSE)
summary(BPRS_ref2)
anova(BPRS_ref2, BPRS_ref1)

```

The plot of BPRSL_my with the observed bprs values will show better the results:

```{r}
p6 <- ggplot(BPRSL_my, aes(x = week, y = bprs, treatment = subject)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "Rated BPRS") +
  theme(legend.position = "top")
print(p6)

```

To finish this analysis exercise, I will create a vector of the fitted values, adding a new column fitted to RATSL and drawing the plot of BPRSL_my with the Fitted values of bprs:

```{r}
Fitted <- fitted(BPRS_ref2)
BPRSL_my <- mutate(BPRSL_my, fitted = Fitted)
p7 <- ggplot(BPRSL_my, aes(x = week, y = fitted, treatment = subject)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "Fitted BPRS") +
  theme(legend.position = "top")
print(p7)
```



